---
title: "time_series_analysis"
output: html_document
---

(SUMMARY HERE)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```

##Time Series Analysis in R
Exploring daily sales data of software firm 1C. This dataset was obtained from Kaggle and is used for the final project of the Coursera "How to win a data science competition" coutse.  NOTE: I did not take this course!

First to import the data and see how it looks:


```{r import}
#train data
train <- read.csv('data/train.csv', stringsAsFactors = FALSE)

#test data & Kaggle sample sub
test <- read.csv('data/test.csv')
sample_sub <- read.csv('data/sample_submission.csv')

#additional data included
items <- read.csv('data/items.csv')
categories <- read.csv('data/item_categories.csv')
shops <- read.csv("data/shops.csv")
```

## Describe the datasets

Initial exploration to check data types, sizes, etc. 

```{r explore}
dim(train)
head(train)

dim(test)
head(test)

head(sample_sub)

```

So, looks as if we're to predict the 'item_cnt_month' based upon 'shop_id' and 'item_id'.. A few things this informs us:

Need to craft a 'daily sales'item_cnt_day' varible in the train set into a monthly sale count to have an appropriate dependant variable for model fitting. This should be done by  
  
  -Aggregate daily sales into monthy, split by shop_id and item_id.

Once this is done, we have two other variables of interest in the train:
  -date :  we'll perform some time-series analysis here
  -price: I'm, sure there is some coreelation between this and monthly sales.
  
So, first step is to create the monthly sales variable in train per shop/item. Then we can explore trends and seasonality in the time-series.

```{r set_dependent}
#First set year variable and edit typeas
train$year <- substr(train$date, 7, 10)
train$year <- as.numeric(train$year)

train$month <- substr(train$date, 4, 5)
train$month <- as.numeric(train$month)

train$date <- as.Date(train$date, "%d.%m.%Y")

sapply(train, class)

train_month <- aggregate(item_cnt_day ~ year + month + date_block_num + shop_id + item_id,
                              data = train, FUN=sum)

names(train_month)[names(train_month) == 'item_cnt_day'] <- 'item_cnt_month'
```

Should have done this first probably :/  lets check for missing values real quick.

```{r}
sapply(train, function(x) sum(is.na(x)))

```

Love Kaggle for that.  Lets divert from the train for a second and add the next block count, year, and month to the test set.  We know it's the next year/month occuring right after the end of the train set, which is october 2015.

```{r}
for (i in 1:nrow(test)){
  test[i,4] = 34
  test[i, 5] = 2015
  test[i, 6] = 11
}

colnames(test)[4:6] <- c('date_block_num', 'year', 'month')

```

So we have a train dataset of 'train_month' and test is updated with necessary information. While I imagine we could work on feature generation or bring in the category/item datasets here to add accuracy to the fit, I'm going to focus on normalizing the time-series first and picking our trend/seasonality. 
```{r}


```

##Season Decomposition
```{r}

```
